Finally, functional class scoring enrichment analysis of Infinium DNA methylation data

### 2023-11-04 manuscript/manuscript.Rmd
writing

---
title: "Finally, functional class scoring enrichment analysis of Infinium DNA methylation data"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
bibliography: gmea.bib
csl: plos-computational-biology.csl
---

<strong>
Mark Ziemann<sup>1,2*</sup>,
Mandhri Abeysooriya<sup>2,3*</sup>,
Anusuiya Bora<sup>2</sup>,
Severin Lamon<sup>3</sup>,
Yen Wong<sup>4</sup>,
Jeffrey Craig<sup>4</sup>
</strong>

@@

Simulations were conducted using the seven FCS methods (LAT, LTT, LAW, LAM, ALT, ALW, ALM) and GSAmeth,
with group sizes of 3-12 and delta values between 0.1 and 0.5, and random gene sets with sizes 20, 50
and 100.
True positives, false positives and false negatives were used to calculate overall precision and recall
at these three gene set sizes (Figure 1).
We found that overall recall was positively associated with the size of the gene sets considered.
At 20 genes per set, GSA had the highest overall performance, with F1 of 0.25, followed by LAW with 0.06.At 50 genes per set, GSA and LAM achieved similar F1 values of ~0.39.
At 100 genes per set, LAM achieved the highest F1 performance of 0.70 followed by LAT with 0.61, while
GSA scored only 0.42.
This result suggests that FCS methods perform better than ORA with gene set sizes >50.

![Figure 1. Precision, recall and F1 for enrichment tests with simulated data.]("path/to/image")

Performance results for GSA and LAM methods with 100 genes per set shows that LAM recall was higher for
most simulations, particularly with lower group sizes.
Taken together, this result suggests that LAM offers improved sensitivity over existing approaches when
the gene sets are larger than 50 members.

 in Figure 2.
This 
As expected, precision and recall values were better when group sizes and delta values were larger.

These data showed a strong relationship between the size of the gene sets and the precision and recall,
regardless of the enrichment test used.
On the whole, precision and recall were better

>>>>

---
title: "Finally, functional class scoring enrichment analysis of Infinium DNA methylation data"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
bibliography: manuscript.bib
csl: epigenetics.csl
---

<strong>
Mark Ziemann<sup>1,2*</sup>,
Mandhri Abeysooriya<sup>2,3</sup>,
Anusuiya Bora<sup>2</sup>,
Severin Lamon<sup>3</sup>,
Yen Wong<sup>4</sup>,
Jeffrey Craig<sup>4</sup>
</strong>

@@

Simulations were conducted using the seven FCS methods (LAT, LTT, LAW, LAM, ALT, ALW, ALM) and GSAmeth,
with group sizes of 3-12 and delta values between 0.1 and 0.5, and random gene sets with sizes 20, 50
and 100.
True positives, false positives and false negatives were used to calculate overall precision and recall
at these three gene set sizes (*Figure 1*).
We found that overall recall was positively associated with the size of the gene sets considered.
At 20 genes per set, GSA had the highest overall performance, with F1 of 0.25, followed by LAW with 0.06.At 50 genes per set, GSA and LAM achieved similar F1 values of ~0.39.
At 100 genes per set, LAM achieved the highest F1 performance of 0.70 followed by LAT with 0.61, while
GSA scored only 0.42.
This result suggests that FCS methods perform better than ORA with gene set sizes >50.

![Figure 1. Precision, recall and F1 for enrichment tests with simulated data.](../figs/fig1.png)

Performance results for GSA and LAM methods with 100 genes per set shows that precision was similar, with most simulations yielding
>95% precision (*Figure 2*).
LAM recall was superior for most simulations, particularly with lower group sizes.
Taken together, this result suggests that LAM offers improved sensitivity over existing approaches when
the gene sets are larger than 50 members.

![Figure 2. Precision, recall and F1 for GSA and LAM enrichment tests over a range of group sizes, delta values and gene set sizes.](../figs/fig2.png)

### Sensitivity of GSA and LAM methods with real cancer data

Next, GSA and LAM methods were applied to compare paired normal and cancer methylation data.

* Chart results with 37 pax.

* Show sensitivity analysis with range of samples from 3 to 30.
